\vspace{5cm} \normalsize \textsc{\NetIDb} \\ [2.0cm] \HRule{0.5pt} \\ \LARGE \textbf{\uppercase{\documenttitle}} \HRule{2pt} \\ [0.5cm] \normalsize \textsc{Sergio Castañeiras Morales\\NIU:1598456\\ }\normalsize \vspace*{5\baselineskip}

### 2025-06-03 Manuscript/Plantilla memòria TFG 2024_25.tex
Lassssst version

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
%----------FONT AND GEOMETRY DEFINITION----------%
\usepackage[a4paper, top=3.0cm, bottom=3.0cm, left = 3.0cm, right = 3.0cm]{geometry}
\voffset = -0.75cm
\usepackage{arev}
\usepackage{helvet}
\renewcommand{\rmdefault}{\sfdefault}
\usepackage[math]{blindtext}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd, esint}
%---------SPACING----------%
\usepackage{setspace}
\setstretch{1.5}
\usepackage[table]{xcolor}

@@

%----------TITLE AND PAGE FORMATTING----------%
\newcommand\NomComplet{Sergio Castañeiras Morales}
\newcommand\NIU{1598456}
\newcommand\NomProfeA{Dr. Miquel Moreno Ferrer}
\newcommand\NomProfeB{Dr. Àngels González Lafont}
\newcommand\CodiTFG{TFG2425\_045}
\newcommand\TitolTFG{AI Application for Azophotoswitches Optimization with Pharmacological Interest}
\newcommand\NomCentre{Departament de Quimica } %cal posar un espai al final
\newcommand\NomGrau{Quimica}
\newcommand\Mes{Juny } %cal espai al final
\newcommand\Any{2025}

@@

\thispagestyle{afteres}
\setcounter{page}{1}
\section*{Resum analític}\par
L'intel·ligència artificial es presenta com una de les revolucions del segle XXI. En particular, el sector de la química computacional està sent profundament sacsejat per aquesta revolució. Aprofitant la inèrcia i l'interès creixent en aquest camp, aquest treball pretén aplicar diferents models d'intel·ligència artificial en l'estudi d'una proteïna d'especial interès per a la nostra salut, la Ciclooxigenasa-2 (COX-2).

La \emph{prostaglandin-endoperoxide sinthasa 2} (PTGS2), també coneguda com COX-2, és una proteïna que, en circumstàncies normals, acostuma a romandre inactiva \cite{Cox2Location}, llevat de la seva expressió durant processos inflamatoris. Així mateix, la manca de retorn a nivells baixos d'expressió després de la inflamació ha estat relacionada amb l'aparició de diferents formes de càncer \cite{DefinitionCOX2CancerDictionary}. Aquest fet ha convertit la COX-2 en objecte d'estudi de nombroses investigacions científiques \cite{Cox2InhibitorsReview}, fet que la fa un punt de partida idoni per al desenvolupament d'algoritmes de \emph{Machine Learning}, ja que disposa d'una gran quantitat de dades per entrenar els models i contrastar els resultats.

L'objectiu principal del projecte és el desenvolupament d'un programari generador d'IAs capaces de predir la concentració d'inhibició al 50\% (IC$_{50}$) per a la COX-2\footnote{En realitat, el programari funciona per a qualsevol proteïna amb entrada a la base de dades de ChEMBL \cite{ChemblDatabase}, malgrat que l'objecte d'estudi és la COX-2.} amb la màxima precisió possible. Per fer-ho, s'extreuen totes les dades de molècules conegudes amb un potencial d'inhibició establert per a la COX-2. Després d'un filtratge configurable per l'usuari, es calculen 5.900 descriptors químics per a cadascuna de les entrades amb el programari AlvaDesk \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Seguidament, una part de les dades s'utilitza per entrenar models de \emph{Random Forest} (RF) \cite{MachineLearningRandomForest}, mentre que la resta es reserva per validar la precisió de les prediccions.

@@

Finalment, els models es fan servir per predir l'IC$_{50}$ de 50 \emph{azophotoswitches} dels quals es tenen dades sobre l'energia lliure d'acoblament proporcionades pel Departament de Química Física de la UAB \cite{UAB_ComputationalChemistry}. L'anàlisi estadístic de les prediccions reflecteix una clara correlació entre ambdues quantitats, fet que reforça la hipòtesi del projecte.
\newpage

\section*{Resumen analítico}
La inteligencia artificial se presenta como una de las revoluciones del siglo XXI. En particular, el sector de la química computacional está siendo profundamente sacudido por esta revolución. Aprovechando la inercia y el interés creciente en este campo, este trabajo pretende aplicar diferentes modelos de inteligencia artificial en el estudio de una proteína de especial interés para nuestra salud, la Ciclooxigenasa-2 (COX-2).

La \emph{prostaglandin-endoperoxide sinthasa 2} (PTGS2), también conocida como COX-2, es una proteína que, en circunstancias normales, suele permanecer inactiva \cite{Cox2Location}, excepto por su expresión durante procesos inflamatorios. Asimismo, la falta de retorno a niveles bajos de expresión después de la inflamación ha sido relacionada con la aparición de diferentes formas de cáncer \cite{DefinitionCOX2CancerDictionary}. Este hecho ha convertido a la COX-2 en objeto de estudio de numerosas investigaciones científicas \cite{Cox2InhibitorsReview}, lo que la convierte en un punto de partida idóneo para el desarrollo de algoritmos de \emph{Machine Learning}, ya que dispone de una gran cantidad de datos para entrenar los modelos y contrastar los resultados.

El objetivo principal del proyecto es el desarrollo de un software generador de IAs capaces de predecir la concentración de inhibición al 50\% (IC$_{50}$) para la COX-2\footnote{En realidad, el software funciona para cualquier proteína con entrada en la base de datos de ChEMBL \cite{ChemblDatabase}, aunque el objeto de estudio es la COX-2.} con la máxima precisión posible. Para ello, se extraen todos los datos de moléculas conocidas con un potencial de inhibición establecido para la COX-2. Tras un filtrado configurable por el usuario, se calculan 5.900 descriptores químicos para cada una de las entradas con el software AlvaDesk \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Seguidamente, una parte de los datos se utiliza para entrenar modelos de \emph{Random Forest} (RF) \cite{MachineLearningRandomForest}, mientras que el resto se reserva para validar la precisión de las predicciones.

@@

Finalmente, los modelos se utilizan para predecir el IC$_{50}$ de 50 \emph{azophotoswitches}, de los cuales se tienen datos sobre la energía libre de acoplamiento proporcionados por el Departamento de Química Física de la UAB \cite{UAB_ComputationalChemistry}. El análisis estadístico de las predicciones refleja una clara correlación entre ambas cantidades, lo que refuerza la hipótesis del proyecto.

@@

\section*{Analytical abstract}
Artificial intelligence is emerging as one of the revolutions of the 21st century. In particular, the field of computational chemistry is being profoundly shaken by this revolution. Taking advantage of the momentum and growing interest in this field, this work aims to apply different artificial intelligence models to the study of a protein of special interest to our health, Cyclooxygenase-2 (COX-2).

The \emph{prostaglandin-endoperoxide synthase 2} (PTGS2), also known as COX-2, is a protein that, under normal circumstances, tends to remain inactive \cite{Cox2Location}, except for its expression during inflammatory processes. Likewise, the failure to return to low expression levels after inflammation has been linked to the onset of various forms of cancer \cite{DefinitionCOX2CancerDictionary}. This fact has made COX-2 the subject of numerous scientific investigations \cite{Cox2InhibitorsReview}, making it an ideal starting point for the development of \emph{Machine Learning} algorithms, as it provides a large amount of data for training models and validating results.

The main objective of the project is to develop software capable of generating AIs that can predict the 50\% inhibition concentration (IC$_{50}$) for COX-2\footnote{Actually, the software works for any protein with an entry in the ChEMBL database \cite{ChemblDatabase}, although the study focuses on COX-2.} with the highest possible accuracy. To achieve this, all known molecular data with an established inhibition potential for COX-2 are extracted. After a user-configurable filtering process, 5,900 chemical descriptors are calculated for each entry using the AlvaDesk software \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Subsequently, part of the data is used to train \emph{Random Forest} (RF) models \cite{MachineLearningRandomForest}, while the rest is reserved to validate the accuracy of the predictions.

@@

Finally, the models are used to predict the IC$_{50}$ of 50 \emph{azophotoswitches}, for which data on the free binding energy have been provided by the Physical Chemistry Unit at UAB \cite{UAB_ComputationalChemistry}. The statistical analysis of the predictions shows a clear correlation between both quantities, which supports the project's hypothesis.

@@

\setcounter{page}{1}
\pagenumbering{arabic}
\section{Introduction}\label{sectionIntroduction}
The impact of \gls{ai} on science has been nothing short of a groundbreaking revolution, with few comparable precedents. The rapid advancements in \gls{ai} have transformed numerous scientific fields\cite{AlphaFold2BiologyAndMedicine}\cite{NationalLibraryOfMedicineGeneralArticle}, including computational chemistry. Today, one of the primary goals of computational chemistry is to predict the properties of unstudied substances while minimizing experimental costs. Traditional approaches in chemistry often rely on complex laboratory techniques, which, while effective, can be time-consuming, expensive, and resource-intensive. In contrast, computational chemistry offers a wide range of methods capable of predicting a molecule’s properties with reasonable accuracy. However, when \gls{ai} comes into play, predictions have demonstrated an almost surgical precision.

@@

\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\schemestart
\scalebox{0.7}{
\chemname{\chemfig[angle increment=30, scale=0.1]{
HO-[1](=[3]O)-[-1]-[1]-[-1]-[1]=[-1]-[-3]-[-1]=[-3]-[7]-[-3]=[7]-[-7]-[7]=[-7]-[3]-[-7]-[7]-[-7]-[7]
}}{Arachidonic acid}
}
	\arrow{->[\gls{cox2}]}
\scalebox{0.7}{
\chemname{\chemfig[angle increment=30]{OH-[5](=[3]O)-[7]-[-3]-[7]-[-3]=[7]-[5]>:[7]*6(-(<[-3.8])-O-O-(<[3.8]C)-(<[-1]=[1]-[-1](<:[-3]OH)-[1]-[-1]-[1]-[-1]-[1])-)
}}{Prostaglandin \chemfig{H_2}}
}
\schemestop
\caption{Reaction catalysed by the COX-2 between arachidonic acid to prostaglandin H$_2$.}
\label{Cox2MainReaction}
\end{figure}

Determining an experimental \gls{ic50} value can be both costly and time-consuming\footnote{Usually this parameter is computed throughout the Cheng Prusoff Equation (Definition (\ref{definitionChengPrusoff})) with experimental data.}. On the other hand, \gls{ai} provides an alternative by offering highly accurate predictions based on existing data, optimising research processes, and accelerating scientific discovery. In this scenario this project aims to implement artificial intelligence in computational chemistry, concretely, using \gls{ai}-based algorithms to predict a drug’s inhibition potential\cite{BookIC50} for a given protein with a relativelly good accuracy \footnote{The accuracy of the ML models is discussed on the Results and Discussion section (\ref{sectionResultsAndDiscussion}).}. To achieve this, we make use of the ChEMBL database\cite{ChemblDatabase}, a vast repository of bioactive molecules with drug-like properties. We extract all known molecular data with a documented inhibition potential for the target protein, creating a comprehensive dataset. The chemical descriptors (Definition (\ref{definitionChemicalDescriptors})) \footnote{All the definitions are stored in the Appendix (\ref{appendixRellevantDefinitions}) \emph{Relevant definitions}. We recommend its consultance for further information.}  of each molecule in the database are then computed using AlvaDesk\cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper} software. Around 5000 descriptors are calculated\cite{DescriptorsBook}, which comprehend from the elemental molecular weight to the complex equipotential electronic surface, providing critical information about each compound’s behaviour. The resulting dataset is subsequently used to train \gls{ai} models, enabling them to predict the inhibition potential of unknown compounds. Finally, we evaluate the reliability of each model by testing it against real experimental data.

@@

The \gls{ai} algorithm used is a in this study is a \gls{ml} model known as the \gls{rf} algorithm\cite{MachineLearningRandomForest}\cite{UAB_ComputationalChemistry}, a powerful ensemble learning method that generates multiple decision trees and combines their outputs to improve prediction accuracy. This approach is particularly well-suited for computational chemistry due to its ability to handle large datasets, manage complex relationships between variables, and reduce overfitting. The Random Forest algorithm operates by constructing numerous random decision trees, each trained on different subsets of the dataset. The final prediction is obtained by averaging the outputs of all trees, ensuring robust and reliable results.

@@

We define a molecule as a \emph{photoswitch} if it contains a bond that undergoes a configurational transformation upon photoexcitation. In particular, we take \emph{azophotoswitches} as a class of photoswitches in which the photo-sensitive bond responsible for isomerization is an azo group (i.e. \ch{N=N}). Azophotoswitches represent the most common and widely studied type of photoswitches due to their robust and tunable photochemical properties. For instance, consider the case of \emph{(cis/trans)-N,1-diphenylmethanimine},
\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\schemestart
\scalebox{0.8}{
	\chemfig[angle increment=30]{*6(-=-(-[1]N=[-1]N-*6(-=-=-=))=-=)}}
	\arrow{<=>[$h\nu$]}
\scalebox{0.8}{
	\chemfig[angle increment=30]{
	N(-[-4]*6(-=-=-=))=[0]N-[-2]*6(-=-=-=)
	}}
\schemestop
	\caption{(cis/trans)-N,1-diphenylmethanimine conversion as an example of an azophotoswitch.}
\label{AzophotoswitchExampleFigure}
\end{figure}
The photo-induced cis/trans isomerization leads to distinct bioactivities for each configuration. This variation arises from the stereochemical constraints required for a molecule to bind to a target protein. Typically, a protein’s binding site has a specific shape, and only molecules whose conformation matches this shape can interact effectively. In the context of an azophotoswitch, one isomer may fit precisely into the binding pocket, leading to strong interactions, while the other may not. Accordingly, we refer to the isomer that interacts most effectively with the protein as the active configuration, and the other as the inactive configuration.\footnote{In the context of this project, interaction refers to inhibition of the protein’s activity.} \footnote{Generally, the trans-isomer is considered the active configuration, while the cis-isomer is considered inactive. However, exceptions exist.}

@@

Among the most promising drug candidates for \gls{cox2} inhibition are Celecoxib (Definition \ref{definitionCelecoxib}) and Rofecoxib (Definition \ref{definitionRofecoxib}).\footnote{These compounds are often referred to as coxib drugs due to the common suffix -coxib.} At the Physical Chemistry Unit of the UAB \cite{UAB_ComputationalChemistry}, researchers are investigating azophotoswitches as potential COX-2 inhibitors, using molecular structures inspired by Celecoxib. Various computational analyses have been conducted, with particular attention to the binding free energy ($\Delta G_{\text{binding}}$).\footnote{Additional results are provided in Appendix \ref{appendixTableOfResults}.}

@@

This project aims to develop a fully functional software tool capable of accurately predicting a molecule’s inhibition potential (\gls{ic50}), leveraging existing data from the ChEMBL database \cite{ChemblDatabase}. The software is thoroughly documented and available in the GitHub repository: \href{https://github.com/SirSergi0/Repository—AI-application-for-azophotoswitches-optimization-with-pharmacological-interest}{\emph{AI Application for Azophotoswitches Optimization with Pharmacological Interest}} \cite{GitHub}\cite{azophotoswitches2025}.

The project specifically focuses on the \gls{cox2} protein. Thanks to the availability of a large \gls{ic50} dataset for this target, the model’s predictive reliability can be effectively validated. The prediction process is powered by an \gls{ai} tool designed to deliver high-quality results, taking into account relevant molecular descriptors. The software also serves to cross-check computational results previously obtained by the Physical Chemistry Unit of the UAB \cite{UAB_ComputationalChemistry}. The application centers on azophotoswitches, molecules of particular pharmacological interest as discussed in Section \ref{sectionIntroduction}. Additionally, a molecular docking study is performed on one of the most promising azophotoswitch candidates.

@@

\newpage
\section{Methodology}\label{sectionMethodology}
The source code is entirely stored in the \href{https://github.com/SirSergi0/Repository---AI-application-for-azophotoswitches-optimization-with-pharmacological-interest}{\emph{AI application for azophotoswitches optimization with pharmacological interest} GitHub repository}\cite{GitHub} and it has been entirely written by Sergio Castañeiras Morales for this project\cite{azophotoswitches2025}. The consultace of this repository is highly recommended in order to comprehended the insights and the workflow of the project. By the time this project is presented, it remains as an open-source repository of the MolBioMed research group \cite{molbiomed}.

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
%----------FONT AND GEOMETRY DEFINITION----------%
%\usepackage[a4paper, top=3.0cm, bottom=3.0cm, left = 3.0cm, right = 3.0cm]{geometry}
\usepackage[a4paper, hmargin=3.0cm, vmargin=3cm]{geometry}
\voffset = -0.75cm
\usepackage{arev}
\usepackage{helvet}
\renewcommand{\rmdefault}{\sfdefault}
\usepackage[math]{blindtext}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd, esint}
%---------SPACING----------%
\usepackage{setspace}
\setstretch{1.5}
\usepackage[table]{xcolor}

@@

%----------TITLE AND PAGE FORMATTING----------%
\newcommand\NomComplet{Sergio Castañeiras Morales}
\newcommand\NIU{1598456}
\newcommand\NomProfeA{Dr. Miquel Moreno Ferrer}
\newcommand\NomProfeB{Dr. Àngels González Lafont}
\newcommand\CodiTFG{TFG2425\_045}
\newcommand\TitolTFG{AI Application for Azophotoswitches Optimization with Pharmacological Interest}
\newcommand\NomCentre{Departament de Quimica } %cal posar un espai al final
\newcommand\NomGrau{Quimica}
\newcommand\Mes{Juny } % cal espai al final
\newcommand\Any{2025}

@@

\thispagestyle{afteres}
\setcounter{page}{1}
\section*{Resum analític}\par
\hspace{0.55cm}L'intel·ligència artificial es presenta com una de les revolucions del segle XXI. En particular, el sector de la química computacional està sent profundament sacsejat per aquesta revolució. Aprofitant la inèrcia i l'interès creixent en aquest camp, aquest treball pretén aplicar diferents models d'intel·ligència artificial en l'estudi d'una proteïna d'especial interès per a la nostra salut, la Ciclooxigenasa-2 (\glstext{cox2}).

La \emph{prostaglandina-endoperoxid sintasa 2} (\glstext{ptgs2}), també coneguda com \glstext{cox2}, és una proteïna que, en circumstàncies normals, acostuma a romandre inactiva \cite{Cox2Location}, llevat de la seva expressió durant processos inflamatoris. Així mateix, la manca de retorn a nivells baixos d'expressió després de la inflamació ha estat relacionada amb l'aparició de diferents formes de càncer \cite{DefinitionCOX2CancerDictionary}. Aquest fet ha convertit la \glstext{cox2} en objecte d'estudi de nombroses investigacions científiques \cite{Cox2InhibitorsReview}, fet que la fa un punt de partida idoni per al desenvolupament d'algoritmes de Machine Learning (\glstext{ml}), ja que disposa d'una gran quantitat de dades per entrenar els models i contrastar els resultats.

L'objectiu principal del projecte és el desenvolupament d'un programari generador d'IAs capaces de predir la concentració d'inhibició al 50\% (\glstext{ic50}) per a la \glstext{cox2}\footnote{En realitat, el programari funciona per a qualsevol proteïna amb entrada a la base de dades de ChEMBL \cite{ChemblDatabase}, malgrat que l'objecte d'estudi és la \glstext{cox2}.} amb la màxima precisió possible. Per fer-ho, s'extreuen totes les dades de molècules conegudes amb un potencial d'inhibició establert per a la \glstext{cox2}. Després d'un filtratge configurable per l'usuari, es calculen 5.900 descriptors químics per a cadascuna de les entrades amb el programari AlvaDesk \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Seguidament, una part de les dades s'utilitza per entrenar models de Random Forest (\glstext{rf}) \cite{MachineLearningRandomForest}, mentre que la resta es reserva per validar la precisió de les prediccions.

@@

Finalment, els models es fan servir per predir l'\glstext{ic50} de 50 \emph{azophotoswitches} dels quals es tenen dades sobre l'energia de Gibbs d'acoblament proporcionades pel grup de recerca MolBioMed \cite{molbiomed}. L'anàlisi estadístic de les prediccions reflecteix una clara correlació entre ambdues quantitats, fet que reforça la hipòtesi del projecte.
\newpage

\section*{Resumen analítico}
\hspace{0.55cm}La inteligencia artificial se presenta como una de las revoluciones del siglo XXI. En particular, el sector de la química computacional está siendo profundamente sacudido por esta revolución. Aprovechando la inercia y el interés creciente en este campo, este trabajo pretende aplicar diferentes modelos de inteligencia artificial en el estudio de una proteína de especial interés para nuestra salud, la Ciclooxigenasa-2 (\glstext{cox2}).

La prostaglandina-endoperóxido sintasa 2 (\glstext{ptgs2}), también conocida como \glstext{cox2}, es una proteína que, en circunstancias normales, suele permanecer inactiva \cite{Cox2Location}, excepto por su expresión durante procesos inflamatorios. Asimismo, la falta de retorno a niveles bajos de expresión después de la inflamación ha sido relacionada con la aparición de diferentes formas de cáncer \cite{DefinitionCOX2CancerDictionary}. Este hecho ha convertido a la \glstext{cox2} en objeto de estudio de numerosas investigaciones científicas \cite{Cox2InhibitorsReview}, lo que la convierte en un punto de partida idóneo para el desarrollo de algoritmos de Machine Learning (\glstext{ml}), ya que dispone de una gran cantidad de datos para entrenar los modelos y contrastar los resultados.

El objetivo principal del proyecto es el desarrollo de un software generador de IAs capaces de predecir la concentración de inhibición al 50\% (\glstext{ic50}) para la \glstext{cox2}\footnote{En realidad, el software funciona para cualquier proteína con entrada en la base de datos de ChEMBL \cite{ChemblDatabase}, aunque el objeto de estudio es la \glstext{cox2}.} con la máxima precisión posible. Para ello, se extraen todos los datos de moléculas conocidas con un potencial de inhibición establecido para la \glstext{cox2}. Tras un filtrado configurable por el usuario, se calculan 5.900 descriptores químicos para cada una de las entradas con el software AlvaDesk \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Seguidamente, una parte de los datos se utiliza para entrenar modelos de Random Forest (\glstext{rf})\cite{MachineLearningRandomForest}, mientras que el resto se reserva para validar la precisión de las predicciones.

@@

Finalmente, los modelos se utilizan para predecir el \glstext{ic50} de 50 \emph{azophotoswitches}, de los cuales se tienen datos sobre la energía libre de acoplamiento proporcionados por el grupo de investigación MolBioMed \cite{molbiomed}. El análisis estadístico de las predicciones refleja una clara correlación entre ambas cantidades, lo que refuerza la hipótesis del proyecto.

@@

\section*{Analytical abstract}
\hspace{0.55cm}Artificial intelligence is emerging as one of the revolutions of the 21st century. In particular, the field of computational chemistry is being profoundly shaken by this revolution. Taking advantage of the momentum and growing interest in this field, this work aims to apply different artificial intelligence models to the study of a protein of special interest to our health, Cyclooxygenase-2 (\glstext{cox2}).

The prostaglandin-endoperoxide synthase 2 (\glstext{ptgs2}), also known as \glstext{cox2}, is a protein that, under normal circumstances, tends to remain inactive \cite{Cox2Location}, except for its expression during inflammatory processes. Likewise, the failure to return to low expression levels after inflammation has been linked to the onset of various forms of cancer \cite{DefinitionCOX2CancerDictionary}. This fact has made \glstext{cox2} the subject of numerous scientific investigations \cite{Cox2InhibitorsReview}, making it an ideal starting point for the development of Machine Learning (\glstext{ml}) algorithms, as it provides a large amount of data for training models and validating results.

The main objective of the project is to develop software capable of generating AIs that can predict the 50\% inhibition concentration (\glstext{ic50}) for \glstext{cox2}\footnote{Actually, the software works for any protein with an entry in the ChEMBL database \cite{ChemblDatabase}, although the study focuses on \glstext{cox2}.} with the highest possible accuracy. To achieve this, all known molecular data with an established inhibition potential for \glstext{cox2} are extracted. After a user-configurable filtering process, 5,900 chemical descriptors are calculated for each entry using the AlvaDesk software \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Subsequently, part of the data is used to train Random Forest (\glstext{rf}) models \cite{MachineLearningRandomForest}, while the rest is reserved to validate the accuracy of the predictions.

@@

Finally, the models are used to predict the \glstext{ic50} of 50 \emph{azophotoswitches}, for which data on the free binding energy have been provided by the MolBioMed research group \cite{molbiomed}. The statistical analysis of the predictions shows a clear correlation between both quantities, which supports the project's hypothesis.

@@

\setcounter{page}{1}
\pagenumbering{arabic}
\section{Introduction}\label{sectionIntroduction}
\hspace{0.55cm}The impact of \gls{ai} on science has been nothing short of a groundbreaking revolution, with few comparable precedents. The rapid advancements in \gls{ai} have transformed numerous scientific fields\cite{AlphaFold2BiologyAndMedicine}\cite{NationalLibraryOfMedicineGeneralArticle}, including computational chemistry. Today, one of the primary goals of computational chemistry is to predict the properties of unstudied substances while minimizing experimental costs. Traditional approaches in chemistry often rely on complex laboratory techniques, which, while effective, can be time-consuming, expensive, and resource-intensive. In contrast, computational chemistry offers a wide range of methods capable of predicting a molecule’s properties with reasonable accuracy. However, when \gls{ai} comes into play, predictions have demonstrated an almost surgical precision.

@@

\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\schemestart
\scalebox{0.7}{
\chemname{\chemfig[angle increment=30, scale=0.1, bond style={line width=1pt}]{
HO-[1](=[3]O)-[-1]-[1]-[-1]-[1]=[-1]-[-3]-[-1]=[-3]-[7]-[-3]=[7]-[-7]-[7]=[-7]-[3]-[-7]-[7]-[-7]-[7]
}}{Arachidonic acid}
}
	\arrow{->[\gls{cox2}]}
\scalebox{0.7}{
\chemname{\chemfig[angle increment=30, bond style={line width=1pt}]{OH-[5](=[3]O)-[7]-[-3]-[7]-[-3]=[7]-[5]>:[7]*6(-(<[-3.8])-O-O-(<[3.8]C)-(<[-1]=[1]-[-1](<:[-3]OH)-[1]-[-1]-[1]-[-1]-[1])-)
}}{Prostaglandin \chemfig{H_2}}
}
\schemestop
\caption{Reaction catalysed by the \gls{cox2} from arachidonic acid to prostaglandin H$_2$.}
\label{Cox2MainReaction}
\end{figure}

Determining an experimental \gls{ic50} value can be both costly and time-consuming\footnote{Usually this parameter is computed throughout the Cheng Prusoff Equation (Definition (\ref{definitionChengPrusoff})) with experimental data.}. On the other hand, \gls{ai} provides an alternative by offering highly accurate predictions based on existing data, optimising research processes, and accelerating scientific discovery. In this scenario this project aims to implement artificial intelligence in computational chemistry, concretely, using \gls{ai}-based algorithms to predict a drug’s inhibition potential\cite{BookIC50} for a given protein with a relativelly good accuracy \footnote{The accuracy of the ML models is discussed on the Results and Discussion section (\ref{sectionResultsAndDiscussion}).}. To achieve this, we make use of the ChEMBL database\cite{ChemblDatabase}, a vast repository of bioactive molecules with drug-like properties. We extract all known molecular data with a documented inhibition potential for the target protein, creating a comprehensive dataset. The chemical descriptors (Definition (\ref{definitionChemicalDescriptors})) \footnote{All the definitions are stored in the Appendix (\ref{appendixRelevantDefinitions}) \emph{Relevant definitions}. We recommend its consultance for further information.}  of each molecule in the database are then computed using AlvaDesk\cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper} software. Around 5000 descriptors are calculated\cite{DescriptorsBook}, which comprehend from the elemental molecular weight to the complex equipotential electronic surface, providing critical information about each compound’s behaviour. The resulting dataset is subsequently used to train \gls{ai} models, enabling them to predict the inhibition potential of unknown compounds. Finally, we evaluate the reliability of each model by testing it against real experimental data.

@@

The \gls{ai} algorithm used is a in this study is a \gls{ml} model known as the \gls{rf} algorithm\cite{MachineLearningRandomForest}, a powerful ensemble learning method that generates multiple decision trees (Definition \ref{definitionDecisionTree}) and combines their outputs to improve prediction accuracy. This approach is particularly well-suited for computational chemistry due to its ability to handle large datasets, manage complex relationships between variables, and reduce overfitting. The Random Forest algorithm operates by constructing numerous random decision trees, each trained on different subsets of the dataset. The final prediction is obtained by averaging the outputs of all trees, ensuring robust and reliable results.

@@

We define a molecule as a \emph{photoswitch} if it contains a bond that undergoes a configurational transformation upon photoexcitation. In particular, we take \emph{azophotoswitches} as a class of photoswitches in which the photo-sensitive bond responsible for isomerization is an azo group (i.e. \ch{N=N}). Azophotoswitches represent the most common and widely studied type of photoswitches due to their robust and tunable photochemical properties. For instance, consider the case of \emph{(cis/trans)-N,1-diphenylmethanimine},
\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\schemestart
\scalebox{0.8}{
	\chemfig[angle increment=30, bond style={line width=1pt}]{*6(-=-(-[1]N=[-1]N-*6(-=-=-=))=-=)}}
	\arrow{<=>[$h\nu$]}
\scalebox{0.8}{
	\chemfig[angle increment=30, bond style={line width=1pt}]{
	N(-[-4]*6(-=-=-=))=[0]N-[-2]*6(-=-=-=)
	}}
\schemestop
	\caption{(cis/trans)-N,1-diphenylmethanimine conversion as an example of an azophotoswitch.}
\label{AzophotoswitchExampleFigure}
\end{figure}
The photo-induced cis/trans isomerization leads to distinct bioactivities for each configuration. This variation arises from the stereochemical constraints required for a molecule to bind to a target protein. Typically, a protein’s binding site has a specific shape, and only molecules whose conformation matches this shape can interact effectively. In the context of an azophotoswitch, one isomer may fit precisely into the binding pocket, leading to strong interactions, while the other may not. Accordingly, we refer to the isomer that interacts most effectively with the protein as the active configuration, and the other as the inactive configuration.\footnote{In the context of this project, interaction refers to inhibition of the protein’s activity.} \footnote{Generally, the cis-isomer is considered the active configuration, while the trans-isomer is considered inactive. However, exceptions exist.}

@@

Among the most promising drug candidates for \gls{cox2} inhibition are Celecoxib (Definition \ref{definitionCelecoxib}) depicted in Scheme (\ref{CelecoxibFigure}) and Rofecoxib (Definition \ref{definitionRofecoxib}).\footnote{Naturally, these compounds are often referred to as \emph{coxib drugs} due to the \gls{cox2}.} At the MolBioMed research group \cite{molbiomed}, researchers are investigating azophotoswitches as potential \gls{cox2} inhibitors, using molecular structures inspired by Celecoxib. Various computational analyses have been conducted, with particular attention to the binding free energy ($\Delta G_{\text{binding}}$).\footnote{Additional results are provided in Appendix \ref{appendixTableOfResults}.}

\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\scalebox{0.8}{
\chemfig[angle increment=30, bond style={line width=1pt}]{
H_2N-[1]S(=[4]O)(=[-2]O)-[1]*6(-=-(-[1]N*5(-N=(-[-1](-[-4]F)(-[-2]F)-[1])-=(-[4.5]*6(-=-(-[4.5])=-=))-))=-=)
}}
\caption{Chemical graph of Celecoxib.}
\label{CelecoxibFigure}
\end{figure}

@@

\hspace{0.55cm}This project aims to develop a fully functional software tool capable of accurately predicting a molecule’s inhibition potential (\gls{ic50}), leveraging existing data from the ChEMBL database \cite{ChemblDatabase}. The software is thoroughly documented and available in the GitHub repository: \href{https://github.com/SirSergi0/Repository—AI-application-for-azophotoswitches-optimization-with-pharmacological-interest}{\emph{AI Application for Azophotoswitches Optimization with Pharmacological Interest}} \cite{GitHub}\cite{azophotoswitches2025}.

The project specifically focuses on the \gls{cox2} protein. Thanks to the availability of a large \gls{ic50} dataset for this target, the model’s predictive reliability can be effectively validated. The prediction process is powered by an \gls{ai} tool designed to deliver high-quality results, taking into account relevant molecular descriptors. The software also serves to cross-check computational results previously obtained by the MolBioMed research group \cite{molbiomed}. The application centres on azophotoswitches, molecules of particular pharmacological interest as discussed in Section \ref{sectionIntroduction}. Additionally, a molecular docking study is performed on one of the most promising azophotoswitch candidates.

@@

\newpage
\section{Methodology}\label{sectionMethodology}
\hspace{0.55cm}The source code is entirely stored in the \href{https://github.com/SirSergi0/Repository---AI-application-for-azophotoswitches-optimization-with-pharmacological-interest}{\emph{AI application for azophotoswitches optimization with pharmacological interest} GitHub repository}\cite{GitHub} and it has been entirely written by Sergio Castañeiras Morales for this project\cite{azophotoswitches2025}. The consultace of this repository is highly recommended in order to comprehended the insights and the workflow of the project. By the time this project is presented, it remains as an open-source repository of the MolBioMed research group \cite{molbiomed}.

The target protein's ID is set at \emph{CHEMBL230} corresponding to the \gls{cox2} ID in the ChEMBL database\cite{ChemblDatabase}. Utilising \emph{requests} python package\cite{PythonPackageRequests} a query URL is sent asking for all molecules with a know \gls{ic50} value (Definition \ref{definitionIC50}) with a limit of 1000 entries per request. The process is iterated until all data is extracted leading a total of 7979 molecules. Hence the data-sheet is processed in pandas dataframes \cite{PythonPackagePandas} (Definition \ref{definitionPandasDataFrames}) and encrypted into binary feather files to optimise reading-writing speed. By removing entries with the same canonical \gls{smiles} (Definition \ref{definitionSMILES})  a total of 5112 molecules remain. Among these entries well known drugs such as Celecoxib (Definition \ref{definitionCelecoxib}), Rofecoxib (Definition \ref{definitionRofecoxib}) or even Ibuprofen can be found. However the \gls{ic50} molecules range is comprehended from $10^{-3}$ to $10^{8}$ nM, a counterproductive range for the AI training procedure. Since we are interested in testing azophotoswitches with presumably low \gls{ic50}, training \gls{ml} models with data of the order of $10^{8}$ or $10^{3}$ nM can be counterproductive since the model might misinterpret the data.\footnote{Typically the range of the azophotoswitches \gls{ic50} is arround the Celecoxib's \gls{ic50}, i.e. arround $120$ nM}. Consecutivelly, a hard-coded range is filtered discarding all molecules outside the given range, for the most part of the analysis this range is set at $[0,200]$ nM \footnote{this \gls{ic50} working range is the standard in these kind of studies \cite{MachineLearningPaper5Lipoxygenase}.} which reduces the dataset to 1438 entries, i.e. molecules.
